In high throughput sequencing all normalizations are wrong but some are often useful

### 2025-03-11 information-supplement.Rmd
submitted version Dec 6

---
title: "Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2
"
shorttitle: "Scale ALDEx2 Supp"
author:
- name: Greg Gloor, Michelle Pistner Nixon, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  BiocStyle::pdf_document: default
package: ALDEx2 1.33.1
abstract: |
    Introduction to scale simulation and FDR correction with ALDEx2.
date: 21 August 2023
link-citations: true
---

```{r HvG_function, echo=F}
H <- function(x.vec){
  x.p <- x.vec/sum(x.vec)
  return(-1 * sum(x.p * log2(x.p)))
}
myfun.H <- function(x){ y <- c(x, 1-x); H(y)}
mf.H <- Vectorize(myfun.H)

@@

# note mean is only used when 
HvG <- function(n.random, n.parts, test='U', mean=30){

@@

  # this will generate data of the 
  # correspondence between H and G in a runif
  # for random number vectors of length
  # 2 to 10000. Does this 5 times
  # N approaches 0 difference as the mean of the N approaches infinity
  for(i in 1:n.parts){
    for(j in 1:n.random){	 
     if(test=='N'){ d <- rnorm(val[i], mean=mean) }
     else if (test == 'U'){ d <- runif(val[i]) * mean }
     else if (test == 'B'){ d <- rbeta(val[i], 0.1, 10) * mean }
     d[d < 0] <- 0
     d <- d/sum(d)
     # mean of log(d), exp = G
     mat.log[i,j] <- G(d)
     # entropy
     mat.h[i,j] <- H(d)
    }
}

@@

HvG.plot <- function(HvG.out){

#plot(rep(val,n.random),mat.log- mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
#  log='x', xlab='length vector', ylab='(-1* G) - H')
plot(HvG.out$mat.log, HvG.out$mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
   xlab='G', ylab='H')

#abline(lm((HvG.out$mat.log[,2]-HvG.out$mat.h[,2]) ~ val), col='red')
abline(lm((HvG.out$mat.h[,2]~HvG.out$mat.log[,2])), col='red')

@@

@@

Information  is a fundamental property of all measured systems. For discrete probability vectors, Shannon defined their information properties and launched the field of information theory for communications. Famously, for any probability vector, Shannon set the total entropy as the inverse of the sum of the probability weighted logarithm of the probabilities. Less well known is that an unweighted average entropy was also defined, and that this turns out to be the inverse of the logarithm of the geometric mean of the probability vector. Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector. Here I show that the information theoretic interpretation can help us understand the scale of a system as defined by Nixon and Silverman, and that this interpretation can help in the interpretation of highly asymmetric datasets. I will show the Asymptotic Equipartition Property of information can be defined as a volume for discrete distributions, and how that volume relates to the scale of a system. Finally, I will show how Shannon's entropy can substitute for the geometric mean in common CoDa operations and how that alters the interpretation of the results.

We can think about  scale  from an information theoretic point of view as a measure of how much information, or total uncertainty, is encoded in a particular sample [@Shannon.48;@Jaynes:2003]. In the geometric interpretation of information theory used in quantum information theory, formally described as  the  Asymptotic Equipartition Property of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution. See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a nice explanation of this.

For notational simplicity assume we have a single discrete random variable to represent a probability distribution with d elements; i.e. \( X =  \mathbf{p}_{i=(1 \dots d )} \). In information theory, the elemental amount of information or surprisal for \(p_i\) is the inverse of the logarithm of the elemental probability, \( -log_2(p_i)\)[@rezaInfoIntro]. This measure is often called self-information.

The total entropy of the system  \( H(X) \) is the weighted sum of the elemental information;

\[
H(X) = - \sum_{i=1}^{d} p_i \log_2 p_i
\]

\( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \). We can also calculate the expected amount of information for each observation in the random variable, and this is the mean of the elemental probability. This measure is also called the sample average of the information [@wilde_2017];

\[
 h(X) = - \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\]

@@

\[
 l2G(X) = \log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\];

>>>>

---
title: "In high throughput sequencing all normalizations are wrong but some are often useful"
shorttitle: "Scale explains normalization"
author:
- name: Greg Gloor, Justin Silverman
  affiliation: Dep't of Biochemistry, University of Western Ontario, Penn State
  email: ggloor@uwo.ca
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
  \usepackage{amsmath}
output:
  pdf_document
---

```{r echo=F, warning=F, message=F,comment=F}
# DESeq2
load('analysis/res.Rda')
library(zCompositions)
#ALDEx2
library(ALDEx2)
data(selex)
load(file="analysis/yst.all.Rda")
load(file="analysis/yst.s.all.Rda")
load(file="analysis/yst.1.all.Rda")
# with gamma = 0.5
load('analysis/x.s.all.Rda')

sig.des <- which(res@listData$padj < 0.05)
sig.ald <- which(yst.all$we.eBH < 0.05)
sig.s.ald <- which(yst.s.all$we.eBH < 0.05)

sig.all <- yst.all$we.eBH < 0.05
sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
sig.s <- yst.s.all$we.eBH < 0.05
sig.1 <- yst.1.all$we.eBH < 0.05

tsc <- read.table("~/Documents/0_git/datasets/transcriptome.tsv", header=T, row.names=1, sep="\t")
```

```{r HvG_function, echo=F}
H <- function(x.vec){
  x.p <- x.vec/sum(x.vec)
  x.p <- x.p[! x.p %in% 0]
  return(-1 * sum(x.p * log2(x.p)))
}
myfun.H <- function(x){ y <- c(x, 1-x); H(y)}
mf.H <- Vectorize(myfun.H)
myfun.2H <- function(x){ y <- c(x, 1-x); 2^H(y)}
mf.2H <- Vectorize(myfun.2H)

@@

# note mean is only used when 
HvG <- function(n.random, n.parts, test='U', mean=30, sd=1){

@@

  # this will generate data of the 
  # correspondence between H and G in a runif
  # for random number vectors of length
  # 2 to 10000. Does this 5 times
  # N approaches 0 difference as the mean of the N approaches infinity
  for(i in 1:n.parts){
    for(j in 1:n.random){	 
     if(test=='N'){ d <- rnorm(val[i], mean=mean, sd=sd) }
     else if (test == 'U'){ d <- runif(val[i]) * mean }
     else if (test == 'B'){ d <- rbeta(val[i], 0.1, 10) * mean }
     d[d < 0] <- 0
     d <- d/sum(d)
     # mean of log(d), exp = G
     mat.log[i,j] <- G(d)
     # entropy
     mat.h[i,j] <- H(d)
      }
  }

@@

HvG.plot <- function(HvG.out, points=FALSE){

#plot(rep(val,n.random),mat.log- mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
#  log='x', xlab='length vector', ylab='(-1* G) - H')
plot(HvG.out$mat.log, HvG.out$mat.h, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
   xlab='G', ylab='H')
if(points==TRUE){
  points(HvG.out$mat.log, HvG.out$mat.h, pch=19, cex=0.5, col=rgb(1,0,0,0.2))
}
#abline(lm((HvG.out$mat.log[,2]-HvG.out$mat.h[,2]) ~ val), col='red')
abline(lm((HvG.out$mat.h[,2]~HvG.out$mat.log[,2])), col='red')

@@

```{r echo=F, eval=F}
prior <- 0:10/10
matG <- matrix(data=NA, ncol=14, nrow=11)
matH <- matrix(data=NA, ncol=14, nrow=11)
for(i in 1:11){
  matG[i,] <- apply(selex+prior[i], 2, G)
  matH[i,] <- apply(selex+prior[i], 2, H)
}

plot(prior, matH[,1]-matH[,8], ylim=c(6,8), pch=19, col=rgb(0,0,1,0.3))
points(prior, matG[,1]-matG[,8], pch=19, col=rgb(1,0,1,0.3))
abline(lm(matG[2:11,1]-matG[2:11,8] ~ prior[2:11]), col='red')
abline(lm(matH[2:11,1]-matH[2:11,8] ~ prior[2:11]), col='blue')
for(i in 2:7){ 
  points(prior, matH[,i]-matH[,i+7], pch=19, col=rgb(0,0,1,0.3)) 
  abline(lm(matH[2:11,i]-matH[2:11,i+7] ~ prior[2:11]), col='blue', lty=i)
}
for(i in 2:7){ 
  points(prior, matG[,i]-matG[,i+7], pch=19, col=rgb(1,0,0,0.3)) 
  abline(lm(matG[2:11,i]-matG[2:11,i+7] ~ prior[2:11]), col='red', lty=i)
}

```
# Scale in high throughput sequencing

High throughput sequencing generates  data where the total number of reads is not informative about the environment. These data are compositional, that is, the data is constrained to a fixed upper limit and can be represented by their corresponding proportions (or probabilities) without losing any information. These data must be normalized and many of the normalizations in widespread use can be represented as ratios, further the normalized data are almost universally log-transformed prior to analysis. Thus, all of the aforementioned normalizations are  actually log-ratios of some sort. There has been much debate in the literature, and much misunderstanding by the user base as to which normalization is the best because until now the actual assumptions of those normalizations could not be understood inside a common framework.

Recently, Silverman and colleagues proved that log-ratios could be understood as  implicit assumptions about the actual scale (size) of the environment. Informally, they recognized that an underlying environment \(W\) containing \(D\) parts in \(N\) samples  could be represented by a count matrix decomposed into proportional (p) and total (t) sub-matrices for each sample \(n\)

\begin{equation}
 W_{dn} = W_{dn}^p \cdot  W_n^t
 \label{envt}
\end{equation}

or the logarithmic form

\begin{equation}
 \log W_{dn} = \log W_{dn}^p  + \log W_n^t,
 \label{logenvt}
\end{equation}

where \begin{equation} W_n^t=\sum{W_{1 \ldots D, n}} \label{sumenvt}\end{equation}  and \begin{equation}W_{dn}^p = \frac{W_{dn}}{ W_n^t} . \label{propenvt}\end{equation}

The logarithmic form in equation 2 is strikingly similar to the logarithmic formula for the centred log-ratio (CLR) transformation used in compositional data analysis (CoDa). Here the observed data matrix  \(Y\) can be transformed by the CLR
\footnote{The formula for the CLR gives the same answer if the matrix  \(Y\) is the raw count values or if its values have been converted to proportions, or indeed by any other linear transformation. This is the scale-invariance property of CoDa in action.}
  by a taking the log-ratio of each part with the sample geometric mean which we will denote as \(h\):

\begin{equation}
 CLR(Y_{dn}) = \log Y_{dn}  + h_n,
 \label{clr}
\end{equation}

where  \begin{equation} h_n = - \frac{1}{D} \sum log Y_{1 \dots D, n}. \label{geometric}\end{equation}

In other words, if \(Y\) is a sample of \(W\), then  \(h_n\) is an implicit estimate of the scale  \( \log W_n^t\). Of course, just because the equations are similar does not guarantee that the values estimated are correct. They went on to show that in the rare instance when the geometric mean was a good estimate of the scale then analyses were trustworthy, but more commonly that the geometric mean was a biased scale estimator of of the data.  Nixon et al made two further important  conclusions: first,  that any estimate that used the logarithm of a ratio approach was also estimating the scale; and second, that the only the relative scale was important. We will discuss the implications of these conclusions in a later section.

All CoDa based approaches are constrained to real vectors  greater than 0

\begin{equation}
X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x > 0, \sum_X = 1.
\end{equation}

Despite this limitation some advantages of the CoDa framework are that it has true distance metrics and it has a solid geometric interpretation. The usual ways of dealing with 0 values in CoDa approaches are to either impute 0 values, to add a prior to all values, or to restrict the analysis to the parts that include only non-0 values. All these methods restrict the generality of CoDa methods and are rightly pointed out as being weaknesses of the approaches.

Information theory is another framework in which proportional vectors can be examined. Information theory  developed as a way to examine the amount of information needed to encode, transmit and decode information. Thus the concept of size was build in from the beginning and this has been one of the most useful advances in all of computation and statistics.  The domain of information theory is subtly different from CoDa in that 0 values are permitted:

\begin{equation}
X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x \ge 0, \sum_X = 1.
\end{equation}

 In information theory, the elemental amount of information or surprisal for \(x_i\) is the inverse of the logarithm of the elemental probability [@rezaInfoIntro]. This measure is often called self-information or the surprisal and we will denote this as \(I\):

\begin{equation}
I(x_i) = \log_2(\frac{1}{x_i}) = -\log_2{x_i}.
\end{equation}

Logarithms of probabilities are additive so that for multiple independent probabilities the entropy \(H\) which is the expected information content of the random variable \(x\) is

\begin{equation}
H= \sum_{X} x_i I(x_i) = -\sum_{X} x_i\log_2(x_i).
\end{equation}

In this formula Shannon defined  \(x_i\log_2(x_i) = 0, x_i=0\) and this effectively removed the 0 constraint when using information theory. Less well known is that a sample average entropy was also defined

\begin{equation}
h = - \frac{1}{D}\sum log_2(x_i),
\end{equation}

and this is simply the geometric mean. Usefully, \(h\) measure converges on \(H\) as the number of measurements from a "typical subset" of the data approaches infinity; this is the Asymptotic Equipartition Property of information as described in [@mackay2003information;@wilde_2017]. As shown in Chapter 4 of Mackay [@mackay2003information] the typical subset is approximately Gaussian and so if both \(h\) and \(H\) are drawn from this distribution they should be identical in the limit. Another way to think about this is that the two measures converge as the distribution of \(p(x_i\) approaches the uniform \(\frac{1}{N}\). Thus the relationship between \(H\) and \(h\) demonstrates independently that the log of the geometric mean can be interpreted as a size of a probability vector.

One the challenges of the scale-based approach has been to identify appropriate base scale values for informed scale models. It has been noted before that \(h\) can be a good or a poor initial estimate. Recall that the actual scale value is not important, but that the ratio between scale values in each group is important (or the difference in log(scale) values). Figure 1 shows that one potential confounder is that the value of \(h\) is very sensitive to small changes in initial assumptions when the data are sparse. For example, one method of treating 0 values is to add a small prior to each value in the matrix. Figure 1 shows how \(H\) and \(h\) vary for values of this prior in the range of 0.01 to 1. Note that entropy is essentially not affected by these prior values, but that  \(h\) can produce a very different initial estimate of the scale with different priors. Another method of treating 0s is to replace the 0 values with a pseudocount, this approach  produces a result that is similar to adding a relatively large prior in sparse data and similar to adding a small prior in non-sparse data. A final method is to impute 0 values using a CoDa-based approach. This method

As shown in Figure 1, one reason for this is that the geometric mean cannot be calculated when one or more of the entries in the probability vector is 0. Thus, Furthermore, the tight relationship between can be exploited

```{r zeroh, echo=F, fig.cap="Plots of H and h with different values added to 0 count parts. The non-sparse plot shows that H and h are linearly related and have a very narrow range over when samples contain no 0 values. The sparse plot shows that H still varies in a small range (although it is about 10-fold larger) but that h now varies has through an approximate 8-fold range when 0 replacement values between 0.1 and 1. The outlier point is a 0 replacement value of 0.01. The difference plot is the mean difference in either h or H between the non-sparse and sparse samples in the dataset. "}

### inpiration can pass function outputs (even nested) as parameters
# ald.ys.h <- aldex(tsc, conditions=y.co, gamma=aldex.makeScaleMatrix(0.5, mu=apply(tsc, 2, function(x)-H(x)), conditions=y.co))
# x.ys.h <- aldex.clr(tsc, conds=y.co, gamma=aldex.makeScaleMatrix(0.5, mu=apply(tsc, 2, function(x)-H(x)), conditions=y.co))

## SELEX
prior <- c(0:10) / 10
p.out <- matrix(data=NA, nrow=14, ncol=6)
prior[1] <- 0.001
for(i in 1:length(prior)){
p.out[i,1] <- -G(selex[,1]+prior[i])
p.out[i,2] <- -G(selex[,8]+prior[i])
p.out[i,3] <- H(selex[,1]+prior[i])
p.out[i,4] <- H(selex[,8]+prior[i])
Gx <- -(apply(selex+prior[i], 2, G))
p.out[i,5] <- mean(Gx[1:7]-Gx[8:14])
Hx <- -(apply(selex+prior[i], 2, H))
p.out[i,6] <- mean(Hx[1:7]-Hx[8:14])
}
selrp1 <- selex 
selrp1[selrp1==0] <- 1
p.out[12,1] <- -G(selrp1[,1])
p.out[12,2] <- -G(selrp1[,8])
p.out[12,3] <- H(selrp1[,1])
p.out[12,4] <- H(selrp1[,8])
Gx <- -(apply(selrp1, 2, G))
p.out[12,5] <- mean(Gx[1:7]-Gx[8:14])
Hx <- -(apply(selrp1, 2, H))
p.out[12,6] <- mean(Hx[1:7]-Hx[8:14])

sel.GBM <- t(cmultRepl(t(selex), label=0, method="GBM"))
p.out[13,1] <- -G(sel.GBM[,1])
p.out[13,2] <- -G(sel.GBM[,8])
p.out[13,3] <- H(sel.GBM[,1])
p.out[13,4] <- H(sel.GBM[,8])
Gx <- -(apply(sel.GBM, 2, G))
p.out[13,5] <- mean(Gx[1:6]-Gx[8:13])
Hx <- -(apply(sel.GBM, 2, H))
p.out[13,6] <- mean(Hx[1:6]-Hx[8:13])

sel.CZM <- t(cmultRepl(t(selex), label=0, method="CZM"))
p.out[14,1] <- -G(sel.CZM[,1])
p.out[14,2] <- -G(sel.CZM[,8])
p.out[14,3] <- H(sel.CZM[,1])
p.out[14,4] <- H(sel.CZM[,8])
Gx <- -(apply(sel.CZM, 2, G))
p.out[14,5] <- mean(Gx[1:6]-Gx[8:13])
Hx <- -(apply(sel.CZM, 2, H))
p.out[14,6] <- mean(Hx[1:6]-Hx[8:13])

cex=c(rep(0.6,11), 1.5,1.5,1.5)
pch=c(rep(0,12), 1,2)
par(mfrow=c(1,3))
plot(p.out[,1], p.out[,3], xlab="h", ylab="H", main="nonsparse", cex=cex, pch=pch)
plot(p.out[,2], p.out[,4], xlab="h", ylab="H", main="sparse", cex=cex, pch=pch)
stripchart(list(p.out[,5], p.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=cex, pch=pch)

#TSCOME
prior <- c(0:10) / 10
t.out <- matrix(data=NA, nrow=12, ncol=7)
colnames(t.out) <- c("h1", "h2", "H1",H2","mh", 'mH','prior')
prior[1] <- 0.001
for(i in 1:length(prior)){
t.out[i,1] <- -G(tsc[,1]+prior[i])
t.out[i,2] <- -G(tsc[,49]+prior[i])
t.out[i,3] <- H(tsc[,1]+prior[i])
t.out[i,4] <- H(tsc[,49]+prior[i])
Gx <- -(apply(tsc+prior[i], 2, G))
t.out[i,5] <- mean(Gx[1:48]-Gx[49:96])
Hx <- -(apply(tsc+prior[i], 2, H))
t.out[i,6] <- mean(Hx[1:48]-Hx[49:96])
t.out[i,7] <- prior[i]
}
selrp1 <- tsc 
selrp1[selrp1==0] <- 1
t.out[12,1] <- -G(selrp1[,1])
t.out[12,2] <- -G(selrp1[,49])
t.out[12,3] <- H(selrp1[,1])
t.out[12,4] <- H(selrp1[,49])
Gx <- -(apply(selrp1, 2, G))
t.out[12,5] <- mean(Gx[1:48]-Gx[49:96])
Hx <- -(apply(selrp1, 2, H))
t.out[12,6] <- mean(Hx[1:48]-Hx[49:96])
t.out[12,7] <- 1

par(mfrow=c(1,3))
plot(t.out[,1], t.out[,3], xlab="h", ylab="H", main="nonsparse", cex=c(rep(0.6,11), 1.5))
plot(t.out[,2], t.out[,4], xlab="h", ylab="H", main="sparse", cex=c(rep(0.6,11), 1.5))
stripchart(list(t.out[,5], t.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=c(rep(0.6,11), 1.5))

### SS

url <- "https://raw.githubusercontent.com/ggloor/datasets/main/singleCell.tsv"
ss <- read.table(url, header=T, row.names=1, sep='\t')
ss <- ss[,c(1:100,1502:1601)]

prior <- c(0:10) / 10
ss.out <- matrix(data=NA, nrow=12, ncol=6)
prior[1] <- 0.001
for(i in 1:length(prior)){
ss.out[i,1] <- -G(ss[,1]+prior[i])
ss.out[i,2] <- -G(ss[,101]+prior[i])
ss.out[i,3] <- H(ss[,1]+prior[i])
ss.out[i,4] <- H(ss[,101]+prior[i])
Gx <- -(apply(ss+prior[i], 2, G))
ss.out[i,5] <- mean(Gx[1:100]-Gx[101:200])
Hx <- -(apply(ss+prior[i], 2, H))
ss.out[i,6] <- mean(Hx[1:100]-Hx[101:200])
}
ssrp1 <- ss 
ssrp1[ssrp1==0] <- 1
ss.out[12,1] <- -G(ssrp1[,1])
ss.out[12,2] <- -G(ssrp1[,101])
ss.out[12,3] <- H(ssrp1[,1])
ss.out[12,4] <- H(ssrp1[,101])
Gx <- -(apply(ssrp1, 2, G))
ss.out[12,5] <- mean(Gx[1:100]-Gx[101:200])
Hx <- -(apply(ssrp1, 2, H))
ss.out[12,6] <- mean(Hx[1:100]-Hx[101:200])

par(mfrow=c(1,3))
plot(ss.out[,1], ss.out[,3], xlab="h", ylab="H", main="nonsparse", cex=c(rep(0.6,11), 1.5))
plot(ss.out[,2], ss.out[,4], xlab="h", ylab="H", main="sparse", cex=c(rep(0.6,11), 1.5))
stripchart(list(ss.out[,5], ss.out[,6]), method="jitter", vertical=TRUE, main="nonsparse - sparse", group.names=c('h diff', 'H diff') ,cex=c(rep(0.6,11), 1.5))

Many such normalizations have been developed and are in widespread use. These include simple proportions, relative log expression (RLE), trimmed mean of M values (TMM), cumulative sum scaling (CSS), LVHA, IQLR and others. In the case of proportions the assumption being made is that the total sum of \(Y_n\) is the same (or at least directly proportional to) the total sum of \(W_n\). All the other normalizations make the assumption that a subset of the parts in \(Y_n\) are the same (or at least directly proportional to) the same subset of parts in \(W_n\). Thus, while the actual details of how the normalization is conducted vary, the end result is always the same: analyses are conducted with a log of a ratio that can be expressed as a function of the observed data

\begin{equation}
N(Y_{dn}) = \log Y_{dn}  + f(\cdot n),
\label{function} 
\end{equation}

where \(\cdot n\) represents the whole or a subset of the parts in sample \(n\).

Returning to equation \eqref{function} we are now in a position to think about  how the concept of scale allows us to understand the assumptions made by different normalizations and to understand how scale normalization supersedes these individual normalizations, including the CLR.

The simplest normalization is the proportion calculated as in equation \eqref{propenvt}. For proportions the values in the output all sum to 1, and so it  is easy to see that the assumption being made is that the scale of all the \(n\) samples is the same. This holds if we replace 0 values, add priors or remove 0 count parts. In all cases the total remains 1 and the assumption holds. In many cases the assumption that the scale of all samples is nearly constant is likely good enough. For example, it likely holds approximately in data taken from closely controlled experiments where only a small perturbation was made.

However, early on it was noted that simple proportions introduced bias into the analyses and so alternative normalizations were developed. These normalizations were explicitly developed to attempt to scale the output data so that it more closely reflected the environment. They did this by making assumptions that appropriate reference parts, or subsets of parts, could be chosen that would allow a good estimate of \(W_d^n\) from \(Y_d^n\). Multiple such normalizations exist but they differ only in how the subset is chosen. For example the RLE as

There are several ways of determining the appropriate denominator for a composition. The most general is to determine a matrix of all possible pairwise log-ratios; this moves the data onto a real space that is completely interpretable and symmetric [@Aitchison:1986]. However, for large datasets such as those generated by high-throughput sequencing or mass spectroscopy of biological samples where there are thousands of parts, the number of possible ratios becomes computationally infeasible. Thus, a number of other approaches have been proposed that involve different ways of choosing appropriate denominators for the log-ratio. The main problem now being that the choice of denominator can affect the interpretation of the result.

The earliest method used by Aitchison was the ALR (additive log ratio), where a presumed invariant part was chosen as the reference for the denominator. The log-ratio was calculated between each part and the reference resulting in a dataset with 1 less entry than the original. The ALR has the advantage of simplicity and ease of interpretation [@aitchison1982;@Greenacre:2021aa]. The primary disadvantage of the ALR is that it is not isometric, that is, the ALR does not by default  recapitulate the geometry of the all-vs-all log-ratio. Greenacre [@Greenacre:2021aa] showed that it was possible to identify a reference for most datasets that was isometric for all practical purposes. It is noteworthy that the ALR has long been used, without naming it such, in fields such as molecular biology and mass spectrometry where it is common to use a single reference as a standard.

Another approach proposed was the CLR (centred log-ratio), where it is presumed that the geometric mean of the parts in each sample is invariant. The CLR is isometric, and captures all of the ratio information of the all-vs-all approach and has the further advantage of being computationally tractable.

Intermediate denominators have also been proposed. For example, the LVHA (low variance, high abundance) log-ratio

choices of reference give slightly different H ( CLR, ALR, hybrid
  - choice of denominator affects interpretation
  - Greenacre

Silverman and colleagues introduced concept of scale which still uses a log-ratio. The actual method of choosing the reference is nearly immaterial, instead it is the relationship between the denominators for each group that determines if the log-ratio is useful. The key finding was that the denominator chosen was a proxy for the underlying size or scale of the environment. The original work by Silverman and colleagues proved this relationship, here we provide an intuitive demonstration based on information theory.

Information theory and CoDa are two ways to deal with compositional data, but information theory is more general. CoDa approaches are applicable to any real vector of numbers

\begin{equation}
X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x > 0, \sum_X = 1.
\end{equation}

Information Theory approaches are subtly different in that here 0 values are permitted:

\begin{equation}
X = [x_1, x_2 \ldots,x_D], x \in \mathbb{R} | x \ge 0, \sum_X = 1.
\end{equation}

 This s in the l to

Scale is related to entropy
  - Shannon's entropy measures size of the data
  - define entropy
  - define size or scale
  - H is insensitive to 0
  - G \(\sim\) H
    - can use diff H as base

@@

Information content is a fundamental property of all measured systems. Shannon defined the information properties of discrete probability vectors in the field of communications and launched the field of information theory. Information theory concerns itself with  how to encode a dataset optimally to minimize its size when transmitted.

The following example is adapted from Chapter 2 of [@wilde_2017]. If we have information encoded in bits--i.e., either a "0" or "1"--we can imagine a uniform encoding scheme for four symbols

\begin{equation}
x  = [A,C,G,T]
\end{equation}

where the underlying probability of observing each symbol is

\begin{equation}
Pr_x=[\frac{1}{2},\frac{1}{4}, \frac{1}{8}, \frac{1}{8}]. 
\end{equation}

Now we can define two encodings; a uniform encoding

\begin{equation}
U_x=[00,01, 10, 11] 
\end{equation}

and a non-uniform encoding

\begin{equation}
N_x=[0,10, 110, 111]. 
\end{equation}

In the uniform encoding the number of bits per symbol is always 2 and in the non-uniform encoding it varies from one to three, with the most common symbol encoded by one bit and the rarest symbols encoded by three bits. Lets now examine the amount of information needed to encode a string of characters that has a similar frequency to the true background. The 11 character string is:

\begin{equation}
aacbdabbaab.
\end{equation}

In the uniform encoding this needs 22 bits because each symbol requires two bits:

\begin{equation}
\sum{U_x} = 5a + 4b + c + d = 10 + 8 + 2 + 2 = 22.
\end{equation}

In the non-uniform encoding this needs 19 bits because the number of bits for each symbol is encoded by a variable number of bits:

\begin{equation}
\sum{N_x}  = 5a + 4b + c + d = 5(1) + 4(2) + 1(3) + 1(3) = 19.
\end{equation}

Going back to our known background frequencies we can calculate the expected number of bits for a given set of symbols and frequencies. For the uniform encoding we get 
\begin{equation}
\frac{1}{2}(2), \frac{1}{4}(2), \frac{1}{8}(2), \frac{1}{8}(2) = \frac{16}{8} = \frac{8}{4} = 2
\end{equation}

which is 2 bits per symbol. While for the variable length encoding we get

\begin{equation}
p_i = \frac{1}{2}(1), \frac{1}{4}(2), \frac{1}{8}(3), \frac{1}{8}(3) = \frac{14}{8} = \frac{7}{4}
\end{equation}

which is smaller.

We are now able to understand the formula for entropy. For notational simplicity assume we have a single discrete random variable to represent a probability distribution with d elements; i.e. \( X =  \mathbf{p}_{i=(1 \dots d )} \). In information theory, the elemental amount of information or surprisal for \(p_i\) is the inverse of the logarithm of the elemental probability [@rezaInfoIntro]. This measure is often called self-information or the surprisal and I will denote this as \(I\):

\begin{equation}
I(X_i) = \log_2(\frac{1}{X_i}) = -\log_2{X_i}.
\end{equation}

Logarithms of probabilities are additive so that for multiple independent probabilities the expected information content of the random variable \(x\) is

\begin{equation}
\sum_{X} p(X_i) I(X_i) = -\sum_{X}  p(X_i)\log_2(X_i).
\end{equation}

And this is the formula for entropy \( H \)! For any encoding of information in a probability vector Shannon showed that \(H\) is the expected information content of that probability vector. We can demonstrate this property  if we remember that the base 2 logarithms of the fractions from equation 2  are -1, -2, -3, -3. The expected entropy to base 2 is thus:

\begin{equation}
H(x) = -\Bigl(\frac{1}{2}(-1) + \frac{1}{4}(-2) + \frac{1}{8}(-3) + \frac{1}{8}(-3) \Bigr) = \frac{7}{4}
\end{equation}

which is precisely the same result as in equation 9 where we  calculated the average bit length of the infinite length message. Thus, entropy  measures the average information encoded by each symbol in a message. One conclusion of the above discussion is that information has a size and that we can measure that size as the average number of bits needed to encode the information.

Less well known is that a sample average entropy was also defined

\begin{equation}
h = - \frac{1}{d}\sum log_2(p),
\end{equation}

and that this measure converges on \(H\) as the number of measurements from a "typical subset" of the data approaches infinity; this is the Asymptotic Equipartition Property of information as described in [@mackay2003information;@wilde_2017]. As shown in Chapter 4 of Mackay [@mackay2003information] the typical subset is approximately Gaussian and so if both \(h\) and \(H\) are drawn from this distribution they should be identical in the limit. Another way to think about this is that the two measures converge as the distribution of \(p_i\) approaches \(\frac{1}{n}\).

```{r asymptote, echo=F, fig.cap="Plot showing that \\(H\\) and \\(h\\) converge as the size of the dataset increases.  "}
set.seed(13)
coeff <- vector()
s.coeff <- vector()

mn <- c(5,10,30,60,100,200,400,800,1000,2000)
std <- c(10,7,5,2,1,.8,.6,.5,.3,.2)
for(i in 1:10){
  X <- HvG(50, 500, test='N', mean=mn[i], sd=1)
  jnk <- lm((X$mat.h[,2]~X$mat.log[,2]))
  coeff[i] <- as.numeric(jnk$coefficients[1])
  X <- HvG(50, 500, test='N', mean=1000, sd=std[i])
  jnk <- lm((X$mat.h[,2]~X$mat.log[,2]))
  s.coeff[i] <- as.numeric(jnk$coefficients[1])
}
par(mfrow=c(1,1))
plot(mn, -coeff, log='y', xlab="vector mean", ylab="H vs. G intercept")
```

Compositional data analysis is concerned with the analysis of compositions; datasets that are strictly positive and that have an arbitrary upper limit. These can always be reduced to a proportion (or probability) without loss of information. Compositional approaches have become popular for the analysis of high throughput sequencing data, and other high dimensional data generated by 'omics platforms. In  compositional data analysis (CoDa)  one popular data transformation is the centred log-ratio (CLR) transform, which is:

\begin{equation}
clr_x = log_2(x_i) - mean(log_2(x) =  log_2(x_i) + h_x,
\end{equation}

and we can see immediately that the second term is equivalent to \(h(x)\). Thus, information theory and compositional data analysis intersect through the geometric mean of a probability vector.

This intersection of information theory and CoDa  can help us understand the scale of a system as defined by Nixon and Silverman which can help in the interpretation of many high throughput sequencing datasets. Since \(H\) can be defined as a volume for discrete distributions it is reasonable to treat \(h\)  treated similarly.

Thus, the use of the geometric mean as a measure of size is acceptable both from the information theoretic and from a CoDa perspective.

Biological context of H - average amount of information in each entry. CLR is diff between sample avg entropy and each i(x)

We can think about  scale  from an information theoretic point of view as a measure of how much information, or total uncertainty, is encoded in a particular sample [@Shannon.48;@Jaynes:2003]. In the geometric interpretation of information theory used in quantum information theory of information [@Cover:1991;@wilde_2017], entropy can be interpreted as the volume occupied by a probability distribution relative to the maximum total entropy. See chapters 4 of the PhD thesis of Lecamwasam [-@ruvi.blog] for a more complete explanation of this. Furthermore, the  Asymptotic Equipartition Property says that the expected value of h(X) for a large number of random variables approaches H(X) as the number of variables approaches infinity [@Cover:1991;@wilde_2017].

The weighted average entropy of the system  \( H(X) \) is the weighted sum of the elemental information;

\begin{equation}
H(X) = - \sum_{i=1}^{d} p_i \log_2 p_i
\end{equation}

\( H(X) \) corresponds to the mean amount of information in each entry that is needed  to reproduce  \( X \). We can also calculate the unweighted expected amount of information for each observation in the random variable, and this is the mean of the elemental probability. This measure is also called the sample average of the information [@wilde_2017];

\begin{equation}
 h(X) = - \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\end{equation}

@@

\begin{equation}
 l2G(X) = \log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\end{equation};
